---
title: "6. Pivot Tables with dplyr"
author: "Onkar Singh"
date: "2/1/2020"
output: html_document
---


# {.tabset}

## 6.1 Summary
Pivot tables are powerful tools in Excel for summarizing data in different ways. We will create these tables using the group_by and summarize functions from the dplyr package (part of the Tidyverse). We will also learn how to format tables and practice creating a reproducible report using RMarkdown and sharing it with GitHub.

Data used in the synthesis section:

+ File name: lobsters.xlsx and lobsters2.xlsx
+ Description: Lobster size, abundance and fishing pressure (Santa Barbara coast)
+ Link: https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=77&revision=newest
+ Citation: Reed D. 2019. SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012. Environmental Data Initiative. [doi](https://doi.org/10.6073/pasta/a593a675d644fdefb736750b291579a0).

### 6.1.1 Objectives
In R, we can use the dplyr package for pivot tables by using 2 functions group_by and summarize together with the pipe operator %>%. We will also continue to emphasize reproducibility in all our analyses.

+ Discuss pivot tables in Excel
+ Introduce group_by() %>% summarize() from the dplyr package
+ Learn mutate() and select() to work column-wise
+ Practice our reproducible workflow with RMarkdown and GitHub


### 6.1.2 Resources
+ dplyr website: dplyr.tidyverse.org
+ [R for Data Science: Transform Chapter by Hadley Wickham & Garrett Grolemund](https://r4ds.had.co.nz/transform.html)
+ [Intro to Pivot Tables I-III videos by Excel Campus](https://youtu.be/g530cnFfk8Y)
+ [Data organization in spreadsheets by Karl Broman & Kara Woo](https://peerj.com/preprints/3183/)


## 6.2 Overview & setup
Wikipedia describes a pivot table as a “table of statistics that summarizes the data of a more extensive table…this summary might include sums, averages, or other statistics, which the pivot table groups together in a meaningful way.”

Aside: Wikipedia also says that “Although pivot table is a generic term, Microsoft trademarked PivotTable in the United States in 1994.”

Pivot tables are a really powerful tool for summarizing data, and we can have similar functionality in R — as well as nicely automating and reporting these tables.

We will first have a look at our data, demo using pivot tables in Excel, and then create reproducible tables in R.

### 6.2.1 View data in Excel
When reading in Excel files (or really any data that isn’t yours), it can be a good idea to open the data and look at it so you know what you’re up against.

Let’s open the lobsters.xlsx data in Excel.

It’s one sheet, and it’s rectangular. In this data set, every row is a unique observation. This is called “uncounted” data; you’ll see there is no row for how many lobsters were seen because each row is an observation, or an “n of 1”.

But also notice that the data doesn’t start until line 5; there are 4 lines of metadata — data about the data that is super important! — that we don’t want to muddy our analyses.

Now your first idea might be to delete these 4 rows from this Excel sheet and save them on another, but we also know that we need to keep the raw data raw. So let’s not touch this data in Excel, we’ll remove these lines in R. Let’s do that first so then we’ll be all set.

### 6.2.2 RMarkdown setup
Let’s start a new RMarkdown file in our repo, at the top-level (where it will be created by default in our Project). I’ll call mine pivot_lobsters.Rmd.

In the setup chunk, let’s attach our libraries and read in our lobster data. In addition to the tidyverse package we will also use the skimr package. You will have to install it, but don’t want it to be installed every time you write your code. The following is a nice convention for having the install instructions available (on the same line) as the library() call.

```{r}
## attach libraries
library(tidyverse)
library(readxl)
library(here)
library(skimr) # install.packages('skimr')
library(kableExtra) # install.packages('kableExtra')
```

We used the read_excel() before, which is the generic function that reads both .xls and .xlsx files. Since we know that this is a .xlsx file, we will demo using the read_xlsx() function.

We can expect that someone in the history of R and especially the history of the readxl package has needed to skip lines at the top of an Excel file before. So let’s look at the help pages ?read_xlsx: there is an argument called skip that we can set to 4 to skip 4 lines.

```{r}
## read in data
lobsters <- read_xlsx(here("data/lobsters.xlsx"), skip=4)
```

Great. We’ve seen this data in Excel so I don’t feel the need to use head() here like we’ve done before, but I do like having a look at summary statistics and classes.


#### 6.2.2.1 skimr::skim
To look at summary statistics we’ve used summary, which is good for numeric columns, but it doesn’t give a lot of useful information for non-numeric data. So it means it wouldn’t tell us how many unique sites there are in this dataset. To have a look there I like using the skimr package:

```{r}
# explore data
skimr::skim(lobsters) 
```

This skimr:: notation is a reminder to me that skim is from the skimr package. It is a nice convention: it’s a reminder to others (especially you!).

skim lets us look more at each variable. Here we can look at our character variables and see that there are 5 unique sites (in the n_unique output). Also, I particularly like looking at missing data. There are 5 missing values in the size_mm variable.

### 6.2.3 Our task
So now we have an idea of our data. But now we have a task: we’ve been asked by a colleague to report about how the average size of lobsters has changed for each site across time.

We will complete this task with R by using the dplyr package for data wrangling, which we will do after demoing how this would do it with pivot tables in Excel.

## 6.4 group_by() %>% summarize()
In R, we can create the functionality of pivot tables with the same logic: we will tell R to group by something and then summarize by something. Visually, it looks like this:

This graphic is from RStudio’s old-school data wrangling cheatsheet; all cheatsheets available from https://rstudio.com/resources/cheatsheets). It’s incredibly powerful to visualize what we are talking about with our data when do do these kinds of operations.

And in code, it looks like this:

data %>%
  group_by() %>% 
  summarize()
`

It reads: “Take the data and then group by something and then summarize by something”.

The pipe operator %>% is a really critical feature of the dplyr package, originally created for the magrittr package. It lets us chain together steps of our data wrangling, enabling us to tell a clear story about our entire data analysis. This is not only a written story to archive what we’ve done, but it will be a reproducible story that can be rerun and remixed. It is not difficult to read as a human, and it is not a series of clicks to remember. Let’s try it out!


### 6.4.1 group_by one variable
Let’s use group_by() %>% summarize() with our lobsters data, just like we did in Excel. We will first group_by year and then summarize by count, using the function n() (in the dplyr package). n() counts the number of times an observation shows up, and since this is uncounted data, this will count each row.

We can say this out loud while we write it: “take the lobsters data and then group_by year and then summarize by count in a new column we’ll call count_by_year”.

```{r}
lobsters %>%
  group_by(year) %>%
  summarize(count_by_year = n())
```
Notice how together, group_by and summarize minimize the amount of information we see. We also saw this with the pivot table. We lose the other columns that aren’t involved here.

Question: What if you don’t group_by first? Let’s try it and discuss what’s going on

```{r}
lobsters %>%
  summarize(count =  n())
```

So if we don’t group_by first, we will get a single summary statistic (sum in this case) for the whole dataset.

Another question: what if we only group_by?

```{r}
lobsters %>%
  group_by(year)
```
R doesn’t summarize our data, but you can see from the output that it is indeed grouped. However, we haven’t done anything to the original data: we are only exploring. We are keeping the raw data raw.

To convince ourselves, let’s now check the lobsters variable. We can do this by clicking on lobsters in the Environment pane in RStudio.

We see that we haven’t changed any of our original data that was stored in this variable. (Just like how the pivot table didn’t affect the raw data on the original sheet).

Aside: You’ll also see that when you click on the variable name in the Environment pane, View(lobsters) shows up in your Console. View() (capital V) is the R function to view any variable in the viewer. So this is something that you can write in your RMarkdown script, although RMarkdown will not be able to knit this view feature into the formatted document. So, if you want include View() in your RMarkdown document you will need to either comment it out #View() or add eval=FALSE to the top of the code chunk so that the full line reads {r, eval=FALSE}.

### 6.4.2 group_by multiple variables
Great. Now let’s summarize by both year and site like we did in the pivot table. We are able to group_by more than one variable. Let’s do this together:

```{r}
lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n())
```
  
We put the site first because that is what we want as an end product. But we could easily have put year first. We saw visually what would happen when we did this in the Pivot Table.

Great.

### 6.4.3 summarize multiple variables
We can summarize multiple variables at a time.

So far we’ve summarized the count of lobster observations. Let’s also calculate the mean and standard deviation. First let’s use the mean() function to calculate the mean. We do this within the same summarize() function, but we can add a new line to make it easier to read. Notice how when you put your curser within the parenthesis and hit return, the indentation will automatically align.

```{r}
lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n(),
            mean_size_mm = mean(size_mm))
```

Aside Command-I will properly indent selected lines.

Great! But this will actually calculate some of the means as NA because one or more values in that year are NA. So we can pass an argument that says to remove NAs first before calculating the average. Let’s do that, and then also calculate the standard deviation with the sd() function

```{r}
lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n(), 
            mean_size_mm = mean(size_mm, na.rm=TRUE), 
            sd_size_mm = sd(size_mm, na.rm=TRUE))
```

So we can make the equivalent of Excel’s pivot table in R with group_by() %>% summarize().

Now we are at the point where we actually want to save this summary information as a variable so we can use it in further analyses and formatting.

So let’s add a variable assignment to that first line:

```{r}
siteyear_summary <- lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n(), 
            mean_size_mm = mean(size_mm, na.rm = TRUE), 
            sd_size_mm = sd(size_mm, na.rm = TRUE))

## inspect our new variable 
siteyear_summary 
```

### 6.4.4 Table formatting with kable()
There are several options for formatting tables in RMarkdown; we’ll show one here from the kableExtra package and learn more about it tomorrow.

It works nicely with the pipe operator, so we can build do this from our new object:

```{r}
## make a table with our new variable
siteyear_summary %>%
  kable()
```
### 6.4.5 R code in-line in RMarkdown
Before we let you try this on your own, let’s go outside of our code chunk and write in Markdown.

I want to demo something that is a really powerful RMarkdown feature that we can already leverage with what we know in R.

Write this in Markdown but replace the # with a backtick (`): “There are #r nrow(lobsters)# total lobsters included in this report.” Let’s knit to see what happens.

I hope you can start to imagine the possibilities. If you wanted to write which year had the most observations, or which site had a decreasing trend, you would be able to.


There are #r nrow(lobsters)# total lobsters included in this report


### 6.4.6 Activity
Build from our analysis and calculate the median lobster size for each site year. Your calculation will use the size_mm variable and function to calculate the median (Hint: ?median)
create and ggsave() a plot.
Then, save, commit, and push your .Rmd, .html, and .png.

Solution (no peeking):

```{r}
siteyear_summary <- lobsters %>%
  group_by(site, year) %>%
  summarize(count_by_siteyear =  n(), 
            mean_size_mm = mean(size_mm, na.rm = TRUE), 
            sd_size_mm = sd(size_mm, na.rm = TRUE), 
            median_size_mm = median(size_mm, na.rm = TRUE))

## a ggplot option:
ggplot(data = siteyear_summary, aes(x = year, y = median_size_mm, color = site)) +
  geom_line() 
```

```{r}
ggsave(here("figures", "lobsters-line.png"))
```


```{r}
## another option:
ggplot(siteyear_summary, aes(x = year, y = median_size_mm)) +
  geom_col() +
  facet_wrap(~site)
```


```{r}
ggsave(here("figures", "lobsters-col.png"))
```

Don’t forget to knit, commit, and push!

Nice work everybody.

## 6.5 Oh no, they sent the wrong data!
Oh no! After all our analyses and everything we’ve done, our colleague just emailed us at 4:30pm on Friday that he sent the wrong data and we need to redo all our analyses with a new .xlsx file: lobsters2.xlsx, not lobsters.xlsx. Aaaaah!

If we were doing this in Excel, this would be a bummer; we’d have to rebuild our pivot table and click through all of our logic again. And then export our figures and save them into our report.

But, since we did it in R, we are much safer. R’s power is not only in analytical power, but in automation and reproducibility.

This means we can go back to the top of our RMarkdown file, and read in this new data file, and then re-knit. We will still need to check that everything outputs correctly, (and that column headers haven’t been renamed), but our first pass will be to update the filename and re-knit:

```{r}
## read in data
lobsters <- read_xlsx(here("data/lobsters2.xlsx"), skip=4)
```

Read this file and run the above codes and the graphs will be automatically updated. 


### 6.5.2 dplyr::count()
Now that we’ve spent time with group_by %>% summarize, there is a shortcut if you only want to summarize by count. This is with a function called count(), and it will group_by your selected variable, count, and then also ungroup. It looks like this:


```{r}
lobsters %>%
  count(site, year)

## This is the same as:
lobsters %>%
  group_by(site, year) %>% 
  summarize(n = n()) %>%
  ungroup()
```


Hey, we could update our RMarkdown text knowing this: There are #r count(lobsters)# total lobsters included in this summary.

There are `r count(lobsters)' total lobsters included in this summary

## 6.6 mutate()

There are a lot of times where you don’t want to summarize your data, but you do want to operate beyond the original data. This is often done by adding a column. We do this with the mutate() function from dplyr. Let’s try this with our original lobsters data. The sizes are in millimeters but let’s say it was important for them to be in meters. We can add a column with this calculation:

```{r}
lobsters %>%
  mutate(size_m = size_mm / 1000)
```

If we want to add a column that has the same value repeated, we can pass it just one value, either a number or a character string (in quotes). And let’s save this as a variable called lobsters_detailed

```{r}
lobsters_detailed <- lobsters %>%
  mutate(size_m = size_mm / 1000, 
         millenia = 2000,
         observer = "Allison Horst")
```


## 6.7 select()

We will end with one final function, select. This is how to choose, retain, and move your data by columns:

Let’s say that we want to present this data finally with only columns for date, site, and size in meters. We would do this:

```{r}
lobsters_detailed %>%
  select(date, site, size_m)
```

One last time, let’s knit, save, commit, and push to GitHub.

6.7.1 END dplyr-pivot-tables session!